  ---
title: "R Notebook"
output: html_notebook
---

**Load required libraries**
```{r Load libraries, message = FALSE}
if (!require("here", quietly = TRUE)) install.packages("here")
if (!require("edgeR", quietly = TRUE)) BiocManager::install("edgeR")
if (!require("sva", quietly = TRUE)) install.packages("sva")
if (!require("ggfortify", quietly = TRUE)) install.packages("ggfortify")
if (!require("scater", quietly = TRUE)) BiocManager::install("scater")
if (!require("scBatch", quietly = TRUE)) devtools::install_github('tengfei-emory/scBatch')
```


**Read input files**
```{r}
#Children
children_gambia18 <- read.delim(here("data", "children", "Children_GAMBIA_2018_swab.tsv"))
children_UK_cur <- read.delim(here("data", "children", "Children_UK-LSMT_curettage.tsv"))
children_UK_swab <- read.delim(here("data", "children", "Children_UK-LSMT_swab.tsv"))

#Adults
adults_UK_ICL_cur <- read.delim(here("data", "adults" ,"Adults_UK-ICL_curettage.tsv"))
adults_UK_ICL_sam <- read.delim(here("data", "adults" ,"Adults_UK-ICL_SAM-strip.tsv"))
adults_laiv1  <- read.delim(here("data", "adults", "LAIV1_baseline_counts_SimonData.txt"))
adults_laiv2  <- read.delim(here("data", "adults", "LAIV2_baseline_counts_SimonData.txt"))

#Elderly
elderly_agesbatch1 <- read.delim(here("data", "adults", "AGES_batch1_gene_symbol.tsv"))

#phenodata file for adults_agesbatch1
phenodata_agesbatch1 = read.delim(here('data', 'phenodata_batch1.tsv'))

#Gene length file to FPKM normalization on scBatch
gene_length <- read.delim(here('data', "GC_lengths.tsv"))
```


"Children_GAMBIA_2018_swab.tsv" and "AGES_batch1_gene_symbol.tsv" only have the Gene Symbol so is necessary to add the Ensembl symbol to merge all datasets. Genes specified by more than one Ensembl symbol (e.g., 5S_rRNA -> 17 Ensembl symbols) were duplicated in both datasets.


**Merges files and create phenodata file**
```{r}
# Select samples samples from batch1 (Elderly data has batch1 AND batch2)
elderly_agesbatch1 = subset(elderly_agesbatch1, 
                           select = c(colnames(elderly_agesbatch1)[1],
                                      as.vector(phenodata_agesbatch1$sample_id)))

# Add Ensembl symbol to elderly_agesbatch1 using IDs from children_UK_swab
elderly_agesbatch1 <- merge(children_UK_swab[,1:2], elderly_agesbatch1, by.x = 'nameofG', by.y = 'Symbol' , all.y= T)

# Add Ensembl symbol to children_gambia18 using IDs from children_UK_swab
children_gambia18 <- merge(children_UK_swab[,1:2], children_gambia18, by = 'nameofG', all.y= T)


adults <- merge(adults_UK_ICL_cur, adults_UK_ICL_sam[,-2], by = 'ID')
adults <- merge(adults, adults_laiv1, by.x = 'ID', by.y = 'X', all = F)
adults <- merge(adults, adults_laiv2, by.x = 'ID', by.y = 'X')
adults <- merge(adults, elderly_agesbatch1[,-1], by = 'ID', all = F)

children <- merge(children_UK_cur, children_UK_swab[,-2], by = 'ID')
children <- merge(children, children_gambia18[,-1], by='ID')

adults.children <- merge(adults, children[,-2], by='ID', all = F)

rownames(adults.children) <- adults.children$ID

#write.table(adults.children, 
#            here("data", "gene_counts_merged_datasets_children_adults_elderly.txt"),
#            sep='\t',quote=F,row.names = F)
```

```{r}
adults.children.cpm <- as.data.frame(edgeR::cpm(adults.children[,-c(1:2)]))

adults.children.cpm <- cbind(adults.children[,1:2],adults.children.cpm)

sexgenes <- t(adults.children.cpm[adults.children.cpm$nameofG %in%
                                    c("RPS4Y1","KDM5D","XIST"),-c(1:2)])

sexgenes.kmeans <- kmeans(sexgenes,2)

out <- data.frame(label = names(sexgenes.kmeans$cluster), sex = sexgenes.kmeans$cluster)
out$sex <-  ifelse(out$sex == 1,'F','M')

write.table(out, here("data","sex_definition.txt"), sep='\t',quote=F,row.names=F)

```

```{r}
#Dataframe with sample info - Phenodata
children_UK_swab_df <- data.frame(label = colnames(children_UK_swab[,-c(1:2)]), Sample  = 'UK_swab', Group = 'Children', timepoint = 'baseline')
children_UK_cur_df <- data.frame(label = colnames(children_UK_cur[,-c(1:2)]), Sample  = 'UK_cur', Group = "Children", timepoint = 'baseline')
children_gambia18_df <- data.frame(label = colnames(children_gambia18[,-c(1:2)]), Sample  = 'Gambia18', Group = "Children", timepoint = 'baseline')
adults_UK_ICL_sam_df <- data.frame(label = colnames(adults_UK_ICL_sam[,-c(1:2)]), Sample  = 'UK_ICL_sam', Group = "Adults", timepoint = 'baseline')
adults_UK_ICL_cur_df <- data.frame(label = colnames(adults_UK_ICL_cur[,-c(1:2)]), Sample  = 'UK_ICL_cur', Group = "Adults", timepoint = 'baseline')
adults_laiv1_df <- data.frame(label = colnames(adults_laiv1[,-1]), Sample  = 'Laiv1', Group = "Adults", timepoint = 'baseline')
adults_laiv2_df <- data.frame(label = colnames(adults_laiv2[,-1]), Sample  = 'Laiv2', Group = "Adults", timepoint = 'baseline')
elderly_agesbatch1_df <- data.frame(label = colnames(elderly_agesbatch1[,-c(1:2)]), 
                                    Sample  = 'ages_batch1', Group = "Elderly", 
                                    timepoint = phenodata_agesbatch1[match(colnames(elderly_agesbatch1[,-c(1:2)]), phenodata_agesbatch1$sample_id),6])

adults.children_df = rbind(children_UK_swab_df, children_UK_cur_df, 
                           children_gambia18_df, adults_UK_ICL_cur_df, 
                           adults_UK_ICL_sam_df, adults_laiv1_df, 
                           adults_laiv2_df, elderly_agesbatch1_df)

#write.table(adults.children_df, here("data", "sample_information.txt"), 
#            sep='\t',quote=F,row.names = F)
adults.children_df <- read.delim(here("data","sample_information.txt"))
adults.children <- read.delim(here("data",
                                   "gene_counts_merged_datasets_adults_children_older.txt"))
fpkm <- read.delim(here("results","FPKM_merged_datasets_children_adults_elderly.txt"))
```

**Batch effect correction with ComBat-seq**
```{r}
###Sample selection
#Select samples collected via curretage
method <- c("UK_ICL_cur","ages_batch1","UK_cur")

adults.children_df <- read.delim(here("data","sample_information.txt"))

adults.children_df <- adults.children_df[adults.children_df$timepoint == "baseline" &
                                           adults.children_df$sampling.method == "curettage",]
adults.children <- read.delim(here("data","gene_counts_merged_datasets_children_adults_elderly.txt"))

rownames(adults.children) <- adults.children$ID
adults.children <- subset(adults.children, select = as.vector(adults.children_df$label))

#Select baseline samples and collected via curretage on phenodata file
adults.children_df <- adults.children_df[adults.children_df$timepoint == "baseline" &
                                           adults.children_df$Sample %in% method,]

#Select baseline samples on gene count file
rownames(adults.children) <- adults.children$ID
adults.childrensub <- subset(adults.children, select = as.vector(adults.children_df$label))

#Drop old factors
adults.children_df$Sample <- factor(adults.children_df$Sample)

#Combat-seq correction
adjusted_counts <- ComBat_seq(as.matrix(adults.children), batch=adults.children_df$Sample)

adjusted_counts$ID <- rownames(adjusted_counts)

adjusted_counts <- merge(gene_length, adjusted_counts, by="row.names", all=F)

adjusted_counts_fpkm <- as.data.frame(calculateFPKM(adjusted_counts[,-c(1:3)], adjusted_counts$Length))

adjusted_counts_tpm <- as.data.frame(calculateTPM(adjusted_counts[,-c(1:3)], adjusted_counts$Length))

rownames(adjusted_counts_fpkm) <- adjusted_counts$Row.names
rownames(adjusted_counts_tpm) <- adjusted_counts$Row.names
#Filter normalized gene counts 
#Keep genes with cpm >= cpm.threshold at least in percentage_samples 
percentage_samples = 0.5
fpkm.threshold = 1
keep = rowSums(adjusted_counts_fpkm > fpkm.threshold) > dim(adults.children_df )[1]*percentage_samples 


adjusted_counts_fpkm = adjusted_counts_fpkm[keep,]

adjusted_counts_fpkm$label <- rownames(adjusted_counts_fpkm)
adjusted_counts_fpkm <- merge(adults.children[,1:2], adjusted_counts_fpkm, by.x='ID',by.y='label',all=F)


write.table(adjusted_counts_fpkm, here('results',
                                       "FPKM_ajusted_children_adults_elderly_Combat-seq.txt"),
            sep='\t',quote=F,row.names = F)

##TPM
percentage_samples = 0.5
tpm.threshold = 1
keep = rowSums(adjusted_counts_tpm > tpm.threshold) > dim(adults.children_df )[1]*percentage_samples 
adjusted_counts_tpm = adjusted_counts_tpm[keep,]

adjusted_counts_tpm$label <- rownames(adjusted_counts_tpm)
adjusted_counts_tpm <- merge(adults.children[,1:2], adjusted_counts_tpm, by.x='ID',by.y='label',all=F)


write.table(adjusted_counts_tpm, here('results',
                                       "TPM_ajusted_children_adults_elderly_Combat-seq.txt"),
            sep='\t',quote=F,row.names = F)

```

# FPKM normalization
```{r FPKM norm}
adults.children <- merge(gene_length, adults.children, by="row.names", all=F)

fpkm <- as.data.frame(calculateFPKM(adults.children[,-c(1:5)], adults.children$Length))

fpkm <- cbind(adults.children[,c(4:5)],fpkm)

write.table(fpkm, here("results","FPKM_merged_datasets_children_adults_elderly.txt"), 
            sep= '\t', quote=F, row.names=F)
```


**Batch effect correction with scBatch**
```{r}

###Sample selection
#Select samples collected via curretage
method <- c("UK_ICL_cur","ages_batch1","UK_cur")

#Select baseline samples and collected via curretage on phenodata file
adults.children_df <- adults.children_df[adults.children_df$timepoint == "baseline" &
                                           adults.children_df$Sample %in% method,]

#Select baseline samples on gene count file
rownames(fpkm) <- fpkm$ID
fpkmsub <- subset(fpkm, select = as.vector(adults.children_df$label))

#Filter by expression
keep <- rowSums(fpkmsub > 1) > ncol(fpkmsub)*0.5
fpkmsub <- fpkmsub[keep,]

adults.children_df$Sample <- factor(adults.children_df$Sample)
batch <- adults.children_df$Sample

correctedD <- QuantNorm(fpkmsub, as.numeric(as.factor(batch)),
                        logdat=F,method='row/column',cor_method='pearson',max=100)

fpkm_adj <-scBatchCpp(c=as.matrix(fpkmsub), d=correctedD, w=diag(ncol(fpkm)), m=1, max=1000,
                             tol=1e-10, step=0.0001, derif=scBatch::derif, verbose=F)

fpkm_adj <- as.data.frame(fpkm_adj)
colnames(fpkm_adj) <- colnames(fpkmsub)
fpkm_adj$ID <- rownames(fpkmsub)
 
out <- merge(fpkm[,1:2], fpkm_adj, by = "ID", all=F)

write.table(out, here("results","FPKM_ajusted_children_adults_elderly_scBatch.txt"),
            sep="\t", quote=F, row.names=F)

t.correctedmatrix <- as.data.frame(t(correctedmatrix))
t.correctedmatrix$label <- rownames(t.correctedmatrix)

t.correctedmatrix <- merge(adults.children_df, t.correctedmatrix, by = 'label')

#PCA
prcom.correctedmatrix <- prcomp(t.correctedmatrix[,-c(1:3)])

autoplot(prcom.correctedmatrix, t.correctedmatrix, shape = 'Group', 
         colour= 'Sample', size=3) +
  scale_color_manual(values= brewer.pal(8,"Set2")) +
  theme_bw(base_size = 15)

correctedmatrix <- as.data.frame(correctedmatrix)
correctedmatrix$ensemblID <- rownames(correctedmatrix)

x <- merge(children_UK_swab[,1:2], correctedmatrix, by.x = 'ID', by.y = 'ensemblID' , all.x= F)

#write.table(x, here('results',
#                    'gene_counts_batch-effect_correction_Adults-Children_UK_Gambia_UK-ICL_AgesBatch1_scBatch_kendall.txt'), sep='\t',quote=F,row.names = F)
```